---
title: "Epi Analysis of Syphilis"
output: pdf_document
date: '2022-'
bibliography: syphilis.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(epitools)
library(dplyr)
library(ggplot2)
library(readxl)
library(stringr)

```

## Syphilis Epi Analysis

An case study into a 1989 Syphilis dataset, the country is unknown but due to the demographic split I assume it would be in a Southern state of America - I am going to use Georgia as the potential state to base all my population based analysis on. 

Race data extracted table 5
Age Sex Race data extracted table 6

```{r}

source('syphilis_functions.R')
df <- read.delim('data/syphilis_3.txt', sep = ',')
head(df)
```


# this is a dataset which has cases of syhpilis over the course of 1989 
# split by sex, race and age
# as I do not know what country it represents I am going to presume it
# represents England in 1989 

## Demographic

```{r}
df %>% group_by(across(Sex)) %>% summarise(percentage= round(n() / nrow(df) * 100))
```
We have a higher percentage of men within our dataset, this could be for a variety of reasons but as per [@Solomon] states that this std has become a concentrated epidemic within MSM during the 1980s. 

reference: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4470884/

Lets look at the age breakdown of our cases split by age
```{r}
age_density <- df %>% group_by(Age,Sex) %>% summarise(density= round(n() / nrow(df),digits=4))
age_density$Age <- factor(age_density$Age, levels = c("<=14" ,"15-19" ,"20-24", "25-29", 
                                     "30-34", "35-44" ,"45-54",">55"))

ggplot(age_density, aes(Age,density )) + 
  geom_col() + facet_wrap(~Sex) + theme_bw() + 
  labs(y='Density\n', x = '\nAge') + 
  ggtitle('Age Density')
```

The male population appears to be affected significantly more from the age of 20 to 44 compared to females where there are significant drops as they get older, dropping circa 2% as they move up the age band.

This could be due to a variety of reasons, this can be dived into deeper by loooking into NATSAL-1, a sex survey which was undertaken in 1991 can give us a potential insight. There were only 3 MSM participants in the survey[@NATSAL1-Sexual-histories] but information was given on their sexual activities and number of partners:

"if you go cruising that could be 7 in one night"[@NATSAL1-Sexual-histories]

Further more subsequent releases of NATSAL show the age range of 25 - 44 to be more sexually active than other age ranges, this is reflected in the histogram above showing how 20 -44 have proportionally more positive cases than outside that range. 

Ethnicitiy representation

```{r}
df %>% group_by(Sex,Race) %>% summarise(counts= n())
```


I will start by doing the basic relative risk of gender catching syphilis


```{r}
demographic_data <- read.csv('data/gender_age_ethnicity_breakdown.csv', skip = 1)
demographic_data$Gender <- tolower(demographic_data$Gender)

sex_total <- demographic_data %>% group_by(Gender) %>% summarise(counts = sum(count))
```

```{r}
gender_s_counts <- df %>% group_by(across(Sex)) %>% summarise(counts= n()) 
gender_s_counts <- as.data.frame(gender_s_counts %>% left_join(data.frame(Sex = c('Female','Male'), Population = c(sex_total[1,2]$counts,sex_total[2,2]$counts))) %>% mutate(Population = Population - counts))
gender_s_counts$counts <- as.numeric(gender_s_counts$counts )
gender_s_counts$Population <- as.numeric(gender_s_counts$Population )

rr_plot_caption <- paste0('Female Population: ',gender_s_counts[1,3],'\nMale Population: ',gender_s_counts[2,3],'\nFemale Cases: ',gender_s_counts[1,2],'\nMale Cases: ',gender_s_counts[2,2])

gender_s_counts <- matrix(c(gender_s_counts[1,3],gender_s_counts[2,3],gender_s_counts[1,2],gender_s_counts[2,2]), nrow=2)

dimnames(gender_s_counts) <- list("Behavior type" =c("Female", "Male"),
                       "Outcome" =  c("Negative Syp", "Positive Syp")
                      )

risk_ratio_male <- riskratio(gender_s_counts)$measure

risk_ratio_male <- data.frame(gender = c('Male'), 
              lower = c(risk_ratio_male[2,2]), 
              est = c(risk_ratio_male[2,1]),  
              upper =c(risk_ratio_male[2,3])
              )
ggplot(risk_ratio_male, aes(gender, est)) + 
  geom_point(size=2) + 
  geom_errorbar(aes(ymin=lower, ymax=upper), width=.1) + 
  geom_hline(yintercept=1) + 
  coord_trans(y='log10') + 
  labs(y='Risk Ratio', x='', caption = rr_plot_caption) + 
  ggtitle('Relative risk of Male catching syphilis against female') + 
  theme_bw()
```

# Estimate Age band Relative Risk
Using the 1990 Age band population estimates

 
```{r}

relative_risk_age(df,  path = 'data/gender_age_ethnicity_breakdown.csv')
```


```{r}
relative_risk_age(df, 'Male',path = 'data/gender_age_ethnicity_breakdown.csv')
```
```{r}
relative_risk_age(df, 'Female',path = 'data/gender_age_ethnicity_breakdown.csv')
```

